Frustratingly easy multi-domain, multilingual translation with latent multi-task group dropout

### 2021-11-04 2022/acl/acl_latex.tex
adding multilingual scores

\begin{document}
\maketitle
\begin{abstract}
Multi-domain and multilingual translation can be cast under multi-task learning in which each domain, each language pair are considered as a task. Sharing most of the parameters of the model causes a catastrophic interference between unrelated tasks. In this work, we propose equivalently grouping the nodes of each layer and dropping a portion of these groups assigning different sparse sub-networks to different task. We demonstrate a significant improvement in multi-domain and multilingual translation. Furthermore, through an ablation study, we show that the learned dropping patterns are similar in related tasks allowing the knowledge transferring between these task.
\end{abstract}

@@

\section{Multi-task group dropout}
We define a group dropout for a domain $d$ as follow
\begin{equation}
\begin{array}{rcl}
\tilde{h}^l &=& h^l \odot r_l^d ,\\
r_l^d \in \{ 0,1 \}^{d_k}, \\
h^{l+1} &=& \text{LAYER}^{l+1}(h^l) ,\\
l & \in & [0,\cdots,L] ,\\
\end{array}
\end{equation}
where $L$ is number of Transformer layers of our NMT model, the binary vector $r_l^d$ is a dropping mask. The dropping mask $r_l^d$ of $l^{th}$ layer and domain $d$ is defined as follow
\begin{equation}
\begin{array}{rcl}
r_l^d(i) &=& \begin{cases}
      1, & \text{if}\ p \times \frac{d_k}{n_p} \leqslant i < (p+1) \times \frac{d_k}{n_p} \\
      & \text{AND} \  m_l^d(p) \text{\small ==} 1 \\
      0, & \text{otherwise},
    \end{cases} \\
p & \in & \{0,\dots,n_p\text{-}1 \}, \\
d & \in & \{0,\dots,n_d\text{-}1 \}, \\
i & \in & \{0,\dots,d_k\text{-}1 \}, \\
m_l^d & \in & \{0,1\}^{n_p} \ \forall d,l \, \\
\mid m_l^d \mid_{L_1} & = & k , \\
\Delta^{n_p}_k & = & \{ m \in \{0,1\}^{n_p} | \mid m \mid_{L_1} = k \},
\end{array}
\end{equation}
where $n_p$ is the number of dropout groups; $n_d$ is the number of domains; $d_k$ is the size of Transformer layers; $k$ is the number of retained groups at each layer; $m_l^d$ is a binary vector indicating which group of the nodes of the $l^{th}$ layer is dropped in the sub-network modeling domain $d$. The group $p$ is retained for domain $d$ if $m_l^d(p)\text{small ==}1$ and is dropped if $m_l^d(p)\text{small ==}0$. We also denote $\Delta^{n_p}_k$ the set of all admissible dropping masks.

@@

In Appendix \ref{appendix:a}, we will prove that $\lim_{\tau \rightarrow 0}\hat{m}_l^d(\tau) = \tilde{m}_l^d$. Furthermore, we will also provide the computation of $\hat{m}_l^d(\tau)$ and prove that $\hat{m}_l^d(\tau)$ is a differentiable function w.r.t $\Phi_l^d$ and its differentiation is computed via implicit differentiation theorem. During the training, we approximate $\tilde{m}_l^d$ by $\hat{m}_l^d(\tau)$ by gradually decaying the hyperparameter $\tau$ to $0$. The gradient of the term $T_1$ w.r.t $\Phi_l^d$ is computed via the chain rule as follows
\begin{equation}
\frac{\partial T_1}{\Phi_l^d} = \frac{\partial T_1}{\hat{m}_l^d(\tau)} \times \frac{\hat{m}_l^d(\tau)}{\Phi_l^d}
\end{equation}
The gradient $\frac{\partial T_1}{\hat{m}_l^d(\tau)}$ is computed via autograd algorithm while $\frac{\hat{m}_l^d(\tau)}{\Phi_l^d}$ is computed via implicit differentiation, that we will explain carefully in Appendix \ref{appendix:a}.
\section{Experimental settings}
\subsection{Baselines}
\subsubsection{Multi-domain, multilingual translation}
In both cases, we examinate the performance of three following MT systems.
\begin{itemize}
	\item Transformer trained with .
	\item Adapter based Transformer.
	\item Transformer with multi-task group dropout.
\end{itemize}
For the sake of brevity, we put the detailed descriptions of these systems in Appendix \ref{appendix:c}.
\subsection{Datasets}
\subsubsection{Multi-domain translation}
We reuse the same data of a recent work of \citet{Pham21revisiting} in multi-domain translation. For the sake of brevity, we leave most of the details of data in Appendix \ref{appendix:c}.
\subsubsection{Multilingual translation}
We evaluate our model on both one-to-many (O2M) and many-to-one (M2O)
translation tasks. We evaluate models on widely used multilingual translation datasets.
\begin{itemize}
	\item TED8-Related. Following the setting of \citet{Wang20balancing}, we use a subset of translations of \citet{qi18when} between English and eight related languages.
	\item TED8-Diverse. The dataset consists of parallel sentences between English and eight diverse languages as in \citet{Wang20balancing}.
\end{itemize}
The detailed description of the datasets is provided in Appendix \ref{appendix:c}.
\section{Results and ablation studies}
\subsection{Multi-domain translation}
According to Table \ref{tab:mdmt}, \system{LaMGD} shows a significant improvement over the generic system \system{Transformer} with zero extra parameters. Moreover, \system{LaMGD} achieve equivalent performance in average compared to \system{Adapter}, which is carrefully finetuned with a approximately 12.4M additional parameters.
\begin{table*}
  \centering
  \begin{tabular}{|p{4cm}|*{7}{r|}} \hline
    Model / Domain & \multicolumn{1}{c|}{\domain{ med}} & \multicolumn{1}{c|}{\domain{ law}} & \multicolumn{1}{c|}{\domain{bank}} & \multicolumn{1}{c|}{\domain{talk}} & \multicolumn{1}{c|}{\domain{ it }} & \multicolumn{1}{c|}{\domain{ rel}} & \multicolumn{1}{c|}{\domain{avg}} \\ \hline 
    \system{Transformer}  \hfill{\footnotesize[65m]} & 40.3 & 59.5 & 49.8 & 36.4 & 49.0 & 80.0  & 52.5\\
    \revision{\system{Adapter}}   \hfill{\footnotesize[+12.4m]}  & 39.5 & 61.0 & 53.1 & 37.5 & 49.6 & 91.0 & 55.28 \\ 
    \system{LaMGD}   \hfill{\footnotesize[+0m]}  & 40.3 & 60.4 & 52.4 & 39.0 & 52.4 & 87.5 & 55.33 \\ 
    \hline
  \end{tabular}
  \caption{Multi-domain translation}
  \label{tab:mdmt}
\end{table*}
\subsection{Multilingual translation}

\subsection{Similarity between dropping masks}
\section{Related work}
\section{Conclusions and outlook}
\bibliography{bibliography}
\bibliographystyle{acl_natbib}
\appendix
\section{Appendix A}
\label{appendix:a}
Proof of limit. Computation of $\hat{m}_l^d(\tau)$. Theorem of implicit differentiation and Proof of differentiability.
\section{Appendix B}
\label{appendix:b}
In this section, we give a simple proof of Inequality \ref{eq:entropy}. In fact, we only need to prove $\mathbb{H} \big[ P(i_1,\cdots,i_k | \Phi_l^d) \big] \geqslant \mathbb{H} \big[ P(i_1 | \Phi_l^d) \big]$. The proof is as follows
\begin{align*}
&\mathbb{H} \big[ P(i_1,\cdots,i_k | \Phi_l^d) \big] = - \mathbb{E}_{i_1,\cdots,i_k | \Phi_l^d} \big[ \text{log}P(i_1,\cdots,i_k | \Phi_l^d) \big] \\
&= -\mathbb{E}_{i_1,\cdots,i_k | \Phi_l^d} \big[ \displaystyle{\sum_{j=2}^k}\text{log}P(i_j | i_1,\cdots,j_{j-1},\Phi_l^d) +  \text{log} P(i_1 | \Phi_l^d) \big] \\
&\geqslant -\mathbb{E}_{i_1,\cdots,i_k | \Phi_l^d} \big[ \text{log} P(i_1 | \Phi_l^d) \big] \\
&= -\mathbb{E}_{i_1 | \Phi_l^d} \big[ \text{log} P(i_1 | \Phi_l^d) \big] = \mathbb{H} \big[ P(i_1 | \Phi_l^d) \big]
\end{align*}
\section{Appendix C}
\label{appendix:c}
\subsection{Multi-domain translation}

@@

>>>>

\begin{document}
\maketitle
\begin{abstract}
Multi-domain and multilingual translation can be cast under multi-task learning in which each domain or language pair is considered a task. Sharing most of the parameters of the model causes a catastrophic interference between unrelated tasks. In this work, we propose equivalently grouping the nodes of each layer and dropping a portion of these groups. Via the task-dependent dropout mechanism, we assign different sparse sub-networks to different tasks. We demonstrate a significant improvement in multi-domain and multilingual translation. Furthermore, through an ablation study, we show that the learned dropping patterns are similar in related tasks allowing the knowledge transferring between these tasks.

\end{abstract}

@@

\section{Multi-task group dropout}
We define a group dropout for a domain $d$ as follow
\begin{equation}
\begin{array}{rcl}
\tilde{h}^l &=& h^l \odot r_l^d ,\\
r_l^d \in \{ 0,1 \}^{d_k}, \\
h^{l+1} &=& \text{LAYER}^{l+1}(h^l) ,\\
l & \in & [0,\cdots,L] ,\\
\end{array}
\end{equation}
where $L$ is number of Transformer layers of our NMT model, the binary vector $r_l^d$ is a dropping mask. The dropping mask $r_l^d$ of $l^{th}$ layer and domain $d$ is defined as follow
\begin{equation}
\begin{array}{rcl}
r_l^d(i) &=& \begin{cases}
      1, & \text{if}\ p \times \frac{d_k}{n_p} \leqslant i < (p+1) \times \frac{d_k}{n_p} \\
      & \text{AND} \  m_l^d(p) \text{\small ==} 1 \\
      0, & \text{otherwise},
    \end{cases} \\
p & \in & \{0,\dots,n_p\text{-}1 \}, \\
d & \in & \{0,\dots,n_d\text{-}1 \}, \\
i & \in & \{0,\dots,d_k\text{-}1 \}, \\
m_l^d & \in & \{0,1\}^{n_p} \ \forall d,l \, \\
\mid m_l^d \mid_{L_1} & = & k , \\
\Delta^{n_p}_k & = & \{ m \in \{0,1\}^{n_p} | \mid m \mid_{L_1} = k \},
\end{array}
\end{equation}
where $n_p$ is the number of dropout groups; $n_d$ is the number of domains; $d_k$ is the size of Transformer layers; $k$ is the number of retained groups at each layer; $m_l^d$ is a binary vector indicating which group of the nodes of the $l^{th}$ layer is dropped in the sub-network modeling domain $d$. The group $p$ is retained for domain $d$ if $m_l^d(p)\text{\small ==}1$ and is dropped if $m_l^d(p)\text{\small ==}0$. We also denote $\Delta^{n_p}_k$ the set of all admissible dropping masks.

@@

In Appendix \ref{appendix:a}, we will prove that $\lim_{\tau \rightarrow 0}\hat{m}_l^d(\tau) = \tilde{m}_l^d$. Furthermore, we will also provide the computation of $\hat{m}_l^d(\tau)$ and prove that $\hat{m}_l^d(\tau)$ is a differentiable function w.r.t $\Phi_l^d$ and its differentiation is computed via implicit differentiation theorem. During the training, we approximate $\tilde{m}_l^d$ by $\hat{m}_l^d(\tau)$ by gradually decaying the hyperparameter $\tau$ to $0$. The gradient of the term $T_1$ w.r.t $\Phi_l^d$ is computed via the chain rule as follows
\begin{equation}
\frac{\partial T_1}{\Phi_l^d} = \frac{\partial T_1}{\hat{m}_l^d(\tau)} \times \frac{\hat{m}_l^d(\tau)}{\Phi_l^d}
\end{equation}
The gradient $\frac{\partial T_1}{\hat{m}_l^d(\tau)}$ is computed via autograd algorithm while $\frac{\hat{m}_l^d(\tau)}{\Phi_l^d}$ is computed via implicit differentiation, that we will explain carefully in Appendix \ref{appendix:a}.
\section{Experimental settings}
\subsection{Baselines}
\subsubsection{Multi-domain, multilingual translation}
In both cases, we examinate the performance of three following MT systems.
\begin{itemize}
	\item Transformer
	\item Adapter based Transformer
	\item Transformer with multi-task group dropout
\end{itemize}
For the sake of brevity, we put the detailed descriptions of these systems in Appendix \ref{appendix:c}.
\subsection{Datasets}
\subsubsection{Multi-domain translation}
We reuse the same data of a recent work of \citet{Pham21revisiting} in multi-domain translation. For the sake of brevity, we leave most of the details of data in Appendix \ref{appendix:c}.
\subsubsection{Multilingual translation}
We evaluate our model on both one-to-many (O2M) and many-to-one (M2O)
translation tasks. We evaluate models on widely used multilingual translation datasets.
\begin{itemize}
	\item TED8-Related. Following the setting of \citet{Wang20balancing}, we use a subset of translations of \citet{qi18when} between English and eight related languages.
	\item TED8-Diverse. The dataset consists of parallel sentences between English and eight diverse languages as in \citet{Wang20balancing}.
\end{itemize}
The detailed description of the datasets is provided in Appendix \ref{appendix:c}.
\section{Results and ablation studies}
\subsection{Multi-domain translation}
According to Table \ref{tab:mdmt}, \system{LaMGD} shows a significant improvement (+2.78) over the generic system \system{Transformer} with zero extra parameters. Moreover, \system{LaMGD} achieve equivalent performance in average compared to \system{Adapter}, which is carrefully finetuned with a approximately 12.4M additional parameters.
\begin{table*}[h!]
  \centering
  \begin{tabular}{|p{4cm}|*{7}{r|}} \hline
    Model / Domain & \multicolumn{1}{c|}{\domain{ med}} & \multicolumn{1}{c|}{\domain{ law}} & \multicolumn{1}{c|}{\domain{bank}} & \multicolumn{1}{c|}{\domain{talk}} & \multicolumn{1}{c|}{\domain{ it }} & \multicolumn{1}{c|}{\domain{ rel}} & \multicolumn{1}{c|}{\domain{avg}} \\ \hline 
    \system{Transformer}  \hfill{\footnotesize[65m]} & 40.3 & 59.5 & 49.8 & 36.4 & 49.0 & 80.0  & 52.5\\
    \revision{\system{Adapter}}   \hfill{\footnotesize[+12.4m]}  & 39.5 & 61.0 & 53.1 & 37.5 & 49.6 & 91.0 & 55.28 \\ 
    \system{LaMGD}   \hfill{\footnotesize[+0m]}  & 40.3 & 60.4 & 52.4 & 39.0 & 52.4 & 87.5 & 55.33 \\ 
    \hline
  \end{tabular}
  \caption{Multi-domain translation}
  \label{tab:mdmt}
\end{table*}
According to Table \ref{tab:multilingual}, \system{LAMGD} shows a boost of 0.42, 0.33, ,0.32 in average over the multilingual \system{Transformer} in O2M-related, M2O-related, O2M-diverse, M2O-diverse respectively. Significant gains are observed in \domain{bel}, \domain{glg} (both direction), \domain{hin} and \domain{bos} (O2M direction).
\begin{table*}[ht]
  \centering
  \begin{tabular}{|p{4cm}|*{9}{r|}} \hline
    O2M-related & \multicolumn{1}{c|}{\domain{aze}} & \multicolumn{1}{c|}{\domain{ bel}} & \multicolumn{1}{c|}{\domain{ces}} & \multicolumn{1}{c|}{\domain{glg}} & \multicolumn{1}{c|}{\domain{por}} & \multicolumn{1}{c|}{\domain{rus}} & \multicolumn{1}{c|}{\domain{slk}} & \multicolumn{1}{c|}{\domain{tur}} & \multicolumn{1}{c|}{\domain{avg}} \\ \hline 
    \system{Transformer}  \hfill{\footnotesize[91.6m]} & 4.80 &7.30&20.80&21.10&39.70&19.80&22.60&15.20&18.91 \\
    \revision{\system{Adapter}}   \hfill{\footnotesize[13m]} & & & & & & & & & \\ 
    \system{LaMGD}  \hfill{\footnotesize[+0m]}  & 5.20&9.40&20.60&22.80&39.60&19.60&22.40&15.00&19.33 \\ 
	\hline
    \hline
    M2O-related & \multicolumn{1}{c|}{\domain{aze}} & \multicolumn{1}{c|}{\domain{ bel}} & \multicolumn{1}{c|}{\domain{ces}} & \multicolumn{1}{c|}{\domain{glg}} & \multicolumn{1}{c|}{\domain{por}} & \multicolumn{1}{c|}{\domain{rus}} & \multicolumn{1}{c|}{\domain{slk}} & \multicolumn{1}{c|}{\domain{tur}} & \multicolumn{1}{c|}{\domain{avg}} \\ \hline 
    \system{Transformer}  \hfill{\footnotesize[67.8m]} &11.40&16.60&28.50&	27.10&43.70&24.60&30.30&25.60&25.98 \\
    \revision{\system{Adapter}}   \hfill{\footnotesize[13m]} & & & & & & & & & \\ 
    \system{LaMGD}   \hfill{\footnotesize[+0m]}  &11.30&17.40&28.60&28.70&	43.70&24.50&30.70&25.60&26.31 \\ 
    \hline
    \hline
    O2M-diverse & \multicolumn{1}{c|}{\domain{bos}} & \multicolumn{1}{c|}{\domain{mar}} & \multicolumn{1}{c|}{\domain{hin}} & \multicolumn{1}{c|}{\domain{mkd}} & \multicolumn{1}{c|}{\domain{ell}} & \multicolumn{1}{c|}{\domain{bul}} & \multicolumn{1}{c|}{\domain{fra}} & \multicolumn{1}{c|}{\domain{kor}} & \multicolumn{1}{c|}{\domain{avg}} \\ \hline 
    \system{Transformer}  \hfill{\footnotesize[96.9m]} & 10.20&4.0&12.70&22.20&31.80&34.00&38.30&8.30&20.19 \\
    \revision{\system{Adapter}}   \hfill{\footnotesize[13m]}  & & & & & & & & & \\ 
    \system{LaMGD}   \hfill{\footnotesize[+0m]} & & & & & & & & & \\
    \hline 
    \hline
    O2M-diverse & \multicolumn{1}{c|}{\domain{bos}} & \multicolumn{1}{c|}{\domain{mar}} & \multicolumn{1}{c|}{\domain{hin}} & \multicolumn{1}{c|}{\domain{mkd}} & \multicolumn{1}{c|}{\domain{ell}} & \multicolumn{1}{c|}{\domain{bul}} & \multicolumn{1}{c|}{\domain{fra}} & \multicolumn{1}{c|}{\domain{kor}} & \multicolumn{1}{c|}{\domain{avg}} \\ \hline 
    \system{Transformer}  \hfill{\footnotesize[70.4m]} &22.40&9.70&20.50&31.80&37.50&38.70&39.80&19.00&27.43 \\
    \revision{\system{Adapter}}   \hfill{\footnotesize[13m]}  & & & & & & & & & \\ 
    \system{LaMGD}   \hfill{\footnotesize[+0m]} &23.50&9.60&21.50&32.20&37.70&38.60&40.00&18.90&27.75 \\
    \hline
  \end{tabular}
  \caption{Multilingual translation}
  \label{tab:multilingual}
\end{table*}
\subsection{Similarity between dropping masks}

\section{Related work}
\section{Conclusions and outlook}
\bibliography{bibliography}
\bibliographystyle{acl_natbib}
\appendix
\section{Appendix A}
\label{appendix:a}
Proof of limit. Computation of $\hat{m}_l^d(\tau)$. Theorem of implicit differentiation and Proof of differentiability.
\section{Appendix B}
\label{appendix:b}
In this section, we give a simple proof of Inequality \ref{eq:entropy}. In fact, we only need to prove $\mathbb{H} \big[ P(i_1,\cdots,i_k | \Phi_l^d) \big] \geqslant \mathbb{H} \big[ P(i_1 | \Phi_l^d) \big]$. The proof is as follows
\begin{align*}
&\mathbb{H} \big[ P(i_1,\cdots,i_k | \Phi_l^d) \big] = - \mathbb{E}_{i_1,\cdots,i_k | \Phi_l^d} \big[ \text{log}P(i_1,\cdots,i_k | \Phi_l^d) \big] \\
&= -\mathbb{E}_{i_1,\cdots,i_k | \Phi_l^d} \big[ \displaystyle{\sum_{j=2}^k}\text{log}P(i_j | i_1,\cdots,j_{j-1},\Phi_l^d) +  \text{log} P(i_1 | \Phi_l^d) \big] \\
&\geqslant -\mathbb{E}_{i_1,\cdots,i_k | \Phi_l^d} \big[ \text{log} P(i_1 | \Phi_l^d) \big] \\
&= -\mathbb{E}_{i_1 | \Phi_l^d} \big[ \text{log} P(i_1 | \Phi_l^d) \big] = \mathbb{H} \big[ P(i_1 | \Phi_l^d) \big]
\end{align*}
\section{Appendix C}
\label{appendix:c}
\subsection{Multi-domain datasets}

@@

\subsection{Multilingual datasets}

\subsection{Multi-domain systems}

\subsection{Multilingual systems}
